---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Preparation of Environment & Packages
```{r, echo=FALSE, output=FALSE}
library(ggplot2)
library(graphics)
library(dplyr)
library(lubridate)
library(knitr)
library(gridExtra)

# fix date bug
Sys.setenv(TZ = "America/Toronto")
```

## Loading and preprocessing the data

###Load the data
```{r}
## read outcome data
raw_data <- read.csv("activity.csv", na.strings = "NA", stringsAsFactors = FALSE)

str(raw_data)
summary( raw_data )
```

### Process/transform the data

The strategy adopted here is essentially to remove lines containing NA.
```{r}
# filter out records with NA
data <- filter(raw_data, !is.na(steps))

# Convert date strings to time_stamps
data <- mutate(data, datetime=as.POSIXct(date) )  #Convert strings as Lubridate datestime format

str(data)
summary( data ) ## 15264 = 17568 - 2304 : validate we removes the correct number of lines
```

## What is mean total number of steps taken per day?

###Calculate the total number of steps taken per day

```{r}
# create dataframe for total number of steps per day 
df <- mutate(data, date_col = as.POSIXct(datetime)) %>%  #convert strings to POSIX Date for
  group_by(date_col) %>%
  summarize(totalSteps = sum(steps))


data_histogram <- data.frame(x = factor(df$date_col), y = df$totalSteps)
ghist1 <- ggplot( data_histogram , aes( y= y) ) +
  geom_histogram() +
  ggtitle("Total Number of steps taken each day") +
  ylab("Total Number of steps taken each day" ) +
  xlab("Occurence count") +
  xlim(c(0,16)) +
  scale_x_continuous(breaks = seq(0,16,2) ) 
plot(ghist1)

```
###Calculate the  mean and median of the total number of steps taken per day

```{r}
# create dataframe for the mean and median of the total number of steps taken per day
mean <-   summarize(df, mean = mean(totalSteps))
median <- summarize(df, mean = median(totalSteps))  

sprintf("mean = %f , median = %f ", mean, median)

```


## What is the average daily activity pattern?

```{r}
# create dataframe for total number of steps per day 
dfIntervals <- mutate(data, date_col = as.POSIXct(datetime)) %>%
  group_by(interval) %>%
  summarize(averageSteps = mean(steps))

g1 <- ggplot( dfIntervals , aes(x=interval,y=averageSteps) ) +
  geom_line() +
  ggtitle(" Average daily activity pattern ") +
  ylab("Average Steps" ) +
  xlab(  "Interval") 
plot(g1)

maxAvg <- dfIntervals[which.max(dfIntervals$averageSteps),]
sprintf("The interval %i contains the maximum average (%f)", maxAvg$interval, maxAvg$averageSteps )

```

## Imputing missing values
###Number of missing values
```{r}
raw_data %>%
  select(everything()) %>%  # replace to your needs
  summarise_all(funs(sum(is.na(.))))
```
```{r}

filledData <- raw_data %>% 
  mutate(datetime=as.POSIXct(date) ) %>%
  group_by(interval) %>%
  mutate(steps= ifelse(is.na(steps), mean(steps, na.rm=TRUE), steps))

summary(filledData)

df2 <- mutate(filledData, date_col = as.POSIXct(datetime)) %>%  #convert strings to POSIX Date for
  group_by(date_col) %>%
  summarize(totalSteps = sum(steps))


data_histogram2 <- data.frame(x = factor(df2$date_col), y = df2$totalSteps)
ghist2 <- ggplot( data_histogram2 , aes( y= y) ) +
  geom_histogram() +
  ggtitle("(Imputating NA)") +
  ylab("Total Number of steps taken each day" ) +
  xlab("Occurence count") +
  xlim(c(0,16)) +
  scale_x_continuous(breaks = seq(0,16,2) ) 


grid.arrange(ghist1, ghist2,ncol=2)


# create dataframe for the mean and median of the total number of steps taken per day
mean <-   summarize(df2, mean = mean(totalSteps))
median <- summarize(df2, mean = median(totalSteps))  

sprintf("mean = %f , median = %f ", mean, median)

```

###Impact on dailynimber of steps 

```{r}
g1 <- ggplot( df , aes(x=date_col,y=totalSteps), color="blue" ) +
  geom_line() +
  geom_point(data=df2,aes(x=date_col,y=totalSteps), color="red") +
  ggtitle(" Total daily activity pattern ") +
  ylab("total daily Steps" ) +
  xlab(  "Date") 
plot(g1)
```

The impact of imputing missing data on the estimates of the total daily number of steps
 is shown on the previous time serie.  The red points correspond to the data in which we 
 assigned the average of the interval to the missing fields.  The black line correspond 
 to the dataset in which we deleted the lines containing NA.  Red points which are not 
 superimposed with the black line are the result of the imputation of missing values. 

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# create dataframe for total number of steps per day 

weekData <- filledData %>% 
  mutate(datetime=as.POSIXct(date) ) %>%
  mutate(day = weekdays(datetime)) %>%
  mutate(weekday = as.factor(ifelse(wday(datetime,week_start=1)< 6, "weekday", "weekend"))) 
 


dfWeekdayIntervals <- weekData %>% filter(weekday == "weekday" ) %>%
  group_by(interval) %>%
  summarise(averageSteps=mean(steps))


dfWeekEndIntervals <- weekData %>% filter( weekday == "weekend" ) %>%
  group_by(interval) %>%
  summarise(averageSteps=mean(steps))


g1 <- ggplot( dfWeekdayIntervals , aes(x=interval,y=averageSteps) ) +
  geom_line() +
  ggtitle("Weekday Average daily activity") +
  ylab("Average Steps" ) +
  xlab(  "Interval") 

g2 <- ggplot( dfWeekEndIntervals , aes(x=interval,y=averageSteps) ) +
  geom_line() +
  ggtitle("Weekend Average daily activity") +
  ylab("Average Steps" ) +
  xlab(  "Interval") 


grid.arrange(g1, g2,ncol=1)

maxAvg <- dfIntervals[which.max(dfIntervals$averageSteps),]
sprintf("The interval %i contains the maximum average (%f)", maxAvg$interval, maxAvg$averageSteps )

```